Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-07 submission/manuscript.Rmd

The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were only bound by KDM5C in EpiLCs (3677 EpiLC only promoters) (Figure 4B). We then performed gene ontology analysis to compare the known functions of genes with KDM5C bound to their promoter in EpiLCs and PNCs. While there were no ontologies significantly enriched for genes only bound by KDM5C in PNCs, gene ontologies for genes shared between PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 4C). Germline-specific ontologies were only enriched in promoters unique to EpiLCs, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) and meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 3C). We then evaluated KDM5C binding around the transcription start site (TSS) of all germline-enriched genes. While KDM5C was bound to a subset of germline gene promoters in EpiLCs, it was not bound to any in PNCs (Figure 4D-E). Together, these results demonstrate KDM5C is significantly enriched at a subset of germline gene promoters in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons.

We then compared KDM5C binding at promoters of RNA-seq DEGs to determine if the germline mRNAs de-repressed in _Kdm5c_-KO EpiLCs and brain are direct targets of KDM5C (Figure 4F). In EpiLCs, KDM5C was bound to about one third of EpiLC-specific and brain-specific germline DEGs (EpiLC only: `r 18/50*100`%, Brain only: `r signif(7/21*100,3)`%). Again, we did not observe any KDM5C binding at germline gene promoters in PNCs, even for brain-specific DEGs (Figure 4H). Some notable differences in KDM5C binding for EpiLC-specific DEGs included _Dazl_ and _Stra8_, two key drivers of germline identity discussed above. Although both mRNAs are expressed in _Kdm5c_-KO EpiLCs, KDM5C is only bound to _Dazl_'s promoter and not _Stra8_'s in EpiLCs (Figure 4F). In contrast to the unique DEGs, 3 out of the 4 genes dysregulated in both the brain and EpiLCs are bound by KDM5C around the TSS (Figure 4G). Altogether, this suggests KDM5C decommssions germline genes in EpiLCs but is no longer recruited to germline gene promoters in neurons. Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C recruitment to their promoters.

## KDM5C promotes _de novo_ DNA methylation at germline genes

>>>>

The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were only bound by KDM5C in EpiLCs (3677 EpiLC only promoters) (Figure 4B). We then performed gene ontology analysis to compare the known functions of genes with KDM5C bound to their promoter in EpiLCs and PNCs. While there were no ontologies significantly enriched for genes only bound by KDM5C in PNCs, gene ontologies for genes shared between PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 4C). Germline-specific ontologies were only enriched in promoters unique to EpiLCs, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) and meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 3C). We then evaluated KDM5C binding around the transcription start site (TSS) of all germline-enriched genes. In EpiLCs, We observed modest KDM5C signal at about half of all germline genes (Figure 4D). Based on our ChIP-seq peak criteria, KDM5C was strongly bound to about 10% of germline gene promoters in EpiLCs (Figure 4E). In condordance with our gene ontology results, we did not observe KDM5C accumiulation at any germline gene promtoers in PNCs (Figure 4D). Together, these results demonstrate KDM5C is significantly enriched at a subset of germline gene promoters in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons.

- read in the homer results txt file above here
- Add in euler KDM5C-bound vs KDM5C-unbound coloring
		- Make sure instances window is 500 bp, change if not

- Germline genes repressed by other histone modifying enzymes have shared genomic features around the TSS to suppress their expression, such as motifs for E2F6/DP1 (E2F) and MGA/MAX (E-box) and CpG islands (CGIs). We asessed transcription factor motifs enriched at germline gene promoters that are either bound or unbound by KDM5C in EpiLCs using HOMER[XXX] (TSS +/- 500 bp, q-value < 0.1). MAX and E2F6 binding sites were significantly enriched at KDM5C-bound germline genes, with about one third of KDM5C-bound promoters contained one or both of these motifs (Figure 4). Contrastingly, KDM5C-unbound germline genes were enriched for RFX transcription factor motifs, a family of transcription factors important for ciliogenesis. The RFX X-box motif is present at XXX% of KDM5C-unbound promoters (Figure 4).

	- Enriched RFX transcription factors included the spermiogenesis regulator RFX2, whose mRNA is also significantly upregulated in _Kdm5c_-KO EpiLCs (Figure 4). 
	 - https://www.nature.com/articles/srep20435
	 - https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4498915/
	 - https://pubmed.ncbi.nlm.nih.gov/32725242/

Finally, we compared KDM5C binding at promoters of RNA-seq DEGs to determine if the germline mRNAs de-repressed in _Kdm5c_-KO EpiLCs and brain are direct targets of KDM5C (Figure 4F). In EpiLCs, KDM5C was bound to about one third of EpiLC-specific and brain-specific germline DEGs (EpiLC only: `r 18/50*100`%, Brain only: `r signif(7/21*100,3)`%). Again, we did not observe any KDM5C binding at germline gene promoters in PNCs, even for brain-specific DEGs (Figure 4H). Some notable differences in KDM5C binding for EpiLC-specific DEGs included _Dazl_ and _Stra8_, two key drivers of germline identity discussed above. Although both mRNAs are expressed in _Kdm5c_-KO EpiLCs, KDM5C is only bound to _Dazl_'s promoter and not _Stra8_'s in EpiLCs (Figure 4F). In contrast to the unique DEGs, 3 out of the 4 genes dysregulated in both the brain and EpiLCs are bound by KDM5C around the TSS (Figure 4G). Altogether, this suggests KDM5C decommssions germline genes in EpiLCs but is no longer recruited to germline gene promoters in neurons. Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C recruitment to their promoters.

## KDM5C promotes _de novo_ DNA methylation at germline genes
- if there are differences, say at germline gene CpG islands
